---
title: "Factors influencing Total Income in the United States"
subtitle: "Research and Statistical Analysis on Factors influencing Total Income in the United States of America"
author: 
  - Wen Han Zhao
thanks: "Code and data are available at: [https://github.com/younazhao/Income-Influencing-Factors.git](https://github.com/younazhao/Income-Influencing-Factors.git)."
date: 1 December 2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(dplyr)
```

```{r}
#| include: false
#| warning: false
#| message: false

data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```


# Introduction
[] for citation 
appendix 
WEIGHT 
The weight (X42001) is a partially design-based
weight constructed at the Federal Reserve using original selection
probabilities and frame information along with aggregate control
totals estimated from the Current Population Survey.  This weight is a
relatively minor revision of the consistent weight series (X42000)
maintained for the SCFs beginning with 1989 (For a detailed discussion
of these weights, see "Consistent Weight Design for the 1989, 1992,
and 1995 SCFs and the Distribution of Wealth," by Arthur B. Kennickell
and R. Louise Woodburn, Review of Income and Wealth, Series 45, Number
2, June 1999, pp. 193-215 or the longer version given on the SCF web
site at www.federalreserve.gov/econresdata/scf/scf_workingpapers.htm).
The nature of the revisions to the consistent weights is described in
"Revisions to the SCF Weighting Methodology: Accounting for
Race/Ethnicity and Homeownership" (Arthur B. Kennickell, January 1999,
www.federalreserve.gov/econresdata/ scf/scf_workingpapers.htm).
A version of the revised weight has been computed for all the surveys
beginning with 1989, and this variable has been added to the public
versions of the SCF data sets.  Users should be aware that the
population defined by the weights for *each implicate* (see above) is
128.6 million households: the sum of each of the weights over all
sample cases and imputation replicates is equal to five times the
number of households in the sample universe.

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (), from .

```{r}
#| label: fig-hist
#| fig-cap: Histogram of Total Income 
#| echo: false
data$log_income <- log(data$income)
ggplot(data, aes(x = log_income)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Total Income", x = "Log Income", y = "Frequency") +
  theme_minimal()
```

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

ggplot(data = data, aes(sample = log_income)) +
  geom_qq(color = "black", size = 1, alpha = 0.7) + # Adds blue points with some transparency
  geom_qq_line(color = "blue", linetype = "dashed", size = 1) + # Adds a red dashed reference line
  theme_minimal() +
  labs(
    title = "Q-Q Plot of Income",
    subtitle = "Comparison of Income Distribution to a Theoretical Normal Distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

```


Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

```{r}
#| label: fig-pred
#| fig-cap: Relationship of Predictors variables with Total Income
#| echo: false
#| warning: false
#| message: false
# Set up the plotting area
par(mfrow = c(3, 3), mar = c(4, 4, 2, 2)) # 3 rows, 3 columns

# Scatterplot: Log Income by Age with a regression line
plot(data$Age, data$log_income, type = "p", main = "Log Income by Age", 
     xlab = "Age", ylab = "Log Income")
abline(lm(log_income ~ Age, data = data), col = "red", lwd = 2) # Add regression line

# List of variables for boxplots
boxplot_vars <- list(
  list(var = "married", title = "Marital Status"),
  list(var = "num_sex", title = "Sex"),
  list(var = "educ", title = "Education Level"),
  list(var = "lf", title = "Labour Force"),
  list(var = "num_work", title = "Work Status"),
  list(var = "num_occu", title = "Occupation")
)

# Generate boxplots in a loop
for (bp in boxplot_vars) {
  boxplot(
    log_income ~ get(bp$var, data), 
    data = data,
    main = paste("Log Income by", bp$title),
    xlab = bp$title,
    ylab = "Log Income",
    col = "lightblue",
    border = "darkblue"
  )
}

```

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

#first_model <-readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

#modelsummary::modelsummary(list("First model" = first_model),statistic = "mad",fmt = 2)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

#pp_check(first_model) +
#  theme_classic() +
#  theme(legend.position = "bottom")

#posterior_vs_prior(first_model) +
#  theme_minimal() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
#  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

#plot(first_model, "trace")

#plot(first_model, "rhat")
```



\newpage


# References


